---
title: "Lecture 09 - PCA and Hierarchical Clustering"
author: "Matt Demelo"
date: "4/30/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1: Preparing Wisconsin cancer data for PCA

```{r}
wisc.df <- read.csv("WisconsinCancer.csv")

head(wisc.df)
```


We remove the first 2 columns since patient id is not useful data, only used for identification, and diagnosis isn't important for us.

```{r}
wisc.data <- as.matrix(wisc.df[ ,3:32])
row.names(wisc.data) <- wisc.df$id
head(wisc.data)
```

Storing diagnosis as a logical vector (1 for M, 0 for B)

```{r}
diagnosis <- as.numeric(wisc.df$diagnosis == 'M')

diagnosis
```

Q1: 30 observations for 569 individuals

```{r}
dim(wisc.data)
```

Q2: 10 variables/features have the _mean suffix.

```{r}
length(grep("_mean",colnames(wisc.data)))

```


Q3: 212 are malignant (357 benign)

```{r}
table(wisc.df$diagnosis)
sum(diagnosis)
```

## Part 2: PCA

We first check to see if we should scale.

```{r}
# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)
```

PCA execute

Since the values in wisc.data are very different between variables, we will scale.

```{r}
wisc.pr <- prcomp(wisc.data, scale = T)
sumwiscpr <- summary(wisc.pr)
sumwiscpr

```

Q4: Scaling is false, PC1 captures 98.2% of variance. Scaling is true, PC1 captures 44.27% of variance.

Q5: When scaled, PC's 1, 2, and 3 are required. When not scaled, only PC1 is required.

Q6: When not scaled, only PC1 is required. When scaled, PC's 1 through 7 are required.


Plotting wisc.pr

```{r}
biplot(wisc.pr)
```

Q7: This plot is pretty messy, and nearly impossible to interpret what is going on. A simpler plot of PC1 vs PC2 is much easier to make sense of.


```{r}
plot(wisc.pr$x,  col = diagnosis + 1, main = "PCA of wisc.pr")
```

This plot shows that PC1 is capturing the variance between benign and malignant samples (since we colored based on our diagnostic data)

Here we calculate the variance captured by each principal component.

```{r}
pr.var <- (wisc.pr$sdev)^2
head(pr.var)
```

Here is the proportion of variance captured by each;

```{r}
pve <- round(100*pr.var/(sum(pr.var)), digits = 2)
pve

```

Now we plot variance explained by each PC (Scree plots):

Here is with a line/dot plot:

```{r}
plot(pve, xlab = "Principal Component", ylab = "Proportion of Variance Explained (%)", main = "PVE of wisc.pr",ylim = c(0,100), typ = "o")
```

And alternatively, with a barplot:

```{r}
barplot(pve, ylab = "Proportion of Variance Explained (%)", 
        main = "PVE of wisc.pr",
        names.arg=paste0("PC",1:length(pve)), las = 2,
        axes = F)
axis(2, at = pve, labels = round(pve,0))
```


Now we plot cuulative proportion of variance explained:

```{r}
plot(cumsum(pve), xlab = "Principal Component", 
     ylab = "Cumulative Proportion of Variance Explained (%)", 
     main = "CumPVE of wisc.pr",ylim = c(0,100), typ = "o")
```

Then we make a side-by-side joining of both the PVE and CumPVE plots.

```{r}
par(mfrow=c(1,2))
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained (%)", 
     main = "PVE of wisc.pr",ylim = c(0,100), typ = "o")
plot(cumsum(pve), xlab = "Principal Component", 
     ylab = "Cumulative Proportion of Variance Explained (%)", 
     main = "CumPVE of wisc.pr",ylim = c(0,100), typ = "o")
```

Now let's shift to using the glory of ggplot2

```{r}
# install.packages("factoextra") 
# Remove the # before the above call to install factoextra
library(factoextra)
```

Now: a beautiful, one line scree plot.

```{r}
fviz_eig(wisc.pr, addlabels = T)
```

### Communicating PCA Results

Q9: concave.points_mean is -0.26085376.

```{r}
load_wisc <- wisc.pr$rotation[,1]

```

Q10 The minimum number of principle components required for understanding 80% of the data: 5 PC's, see below in the cumsum(pve) call.

```{r}
cumsum(pve)

cumsum(pve) >= 80
```


## Part 3: Hierarchical clustering.

First we scale our data.

```{r}
data.scaled <- scale(wisc.data)
head(data.scaled)
```

First we calculate Euclidean distances between observations:

```{r}
data.dist = dist(data.scaled)
head(data.dist)
```

Next, we cluster with hclust()

```{r}
wisc.hclust <- hclust(data.dist, "complete")
wisc.hclust
```

### Hierarchical Clustering: Results

Q11:

```{r}
plot(wisc.hclust)
abline(h = 19, col = "red", lty = 2)
```

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)

table(wisc.hclust.clusters)
```

Use the table() function to compare cluster emmbership to actual diagnosis:

```{r}
table(wisc.hclust.clusters, diagnosis)
```


## Part 5: Combination Methods

Combining our PCA with Hclust.

```{r}
# cumsum(pve) >= 90
wisc.pr_90 <- wisc.pr$x[1:7]
wisc.pca.hclust <-  hclust(dist(wisc.pr$x), method = "ward.D2")

plot(wisc.pca.hclust)

```

This gives us a much more legible dendogram which still represents the two major groups in the data seen in the original data.

Let's see how many clusters exist in each:


```{r}
gprs <- cutree(wisc.pca.hclust,k = 2)

table(gprs)
```

```{r}
table(gprs,diagnosis)
```

Now we can do some PCA plotting using the data obtained from hclust.

```{r}
plot(wisc.pr$x[,1:2], col = gprs)
```


```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], col = (diagnosis + 1))
```

#### Now to look at our data in 3D using the rgl package:

```{r}
#install.packages("rgl") # remove the "#" to actually install this.
library(rgl)

# ** WARNING ** xquarts package must be installed or this won't work properly.


plot3d(wisc.pr$x[,1:3], 
       xlab = "PC1", ylab = "PC2", zlab = "PC3",
       main = "PCA Plot of Wisconsin Cancer Data",
       cex = 1.5, typ = "s", size  = 1, 
       col = (diagnosis + 1))
rglwidget(width = 400, height = 400)
```


Now we look at a "cleaner" clustering:

```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:7]), method = "ward.D2")
wisc.pr.hclust.clusters <-  cutree(wisc.pr.hclust, k =2)
table(wisc.pr.hclust.clusters, diagnosis)
```

## Part 7: Prediction

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

Now we plot our PCA:

```{r}
plot(wisc.pr$x[,1:2], col=gprs)
points(npc[,1], npc[,2], col="blue", pch=16)
```
Here we have a PCA plot where two specific patients are highlighted, NPC1 and NPC2.
The patient on the right should be prioritized since they fall within the "malignant" colored data cluster.

## Part 8: PCA for Protein Structure Data

http://thegrantlab.org/bio3d/webapps








