---
title: 'Lecture07: R Functions and Packages'
author: "Matt Demelo"
date: "4/23/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Functions Revisited

We will source a file from online with our functions from last lecture (06).

```{r}
source("http://tinyurl.com/rescale-R")
```

Try out the last day's rescale() function:

```{r}
rescale(1:10)
```

```{r}
# rescale2(c(2,3,"String"))
```
This will give an error.


# Find missing NA values in 2 vectors:

A function to count the number of missing value/"NA"s occuring in the same spot in two vectors

```{r}
x <- c(1, 2, NA, 3, NA)
y <- c(NA, 3, NA, 3, 4)

# The answer would be 1 here, but we want R to figure that out for us.
```

The function is.na() will tell us whether a variable is an NA

```{r}
is.na(x)
```

```{r}
is.na(y)
```


Try putting these together with an AND statement, and we get a boolean vector

```{r}
is.na(x) & is.na(y)
```

Take the sum() of the boolean vector to get the number of TRUEs (same as 1) which is the number of positions where the value is NA in both vectors.

```{r}
sum(is.na(x) & is.na(y))
```

How could this be adapted into a function?

```{r}
both_na <- function(vec1, vec2) {
  sum(is.na(vec1) & is.na(vec2))
}
```

Here is the functional output with an example.

```{r}
both_na(x,c(NA,3,NA,2,NA))
```

```{r}
xtest <- c(NA, NA, NA)
y1 <- c(1, NA, NA)
y2 <- c(1, NA, NA, NA)
```

```{r}
both_na(xtest,y2)
```

How did this happen? Let's check a nother case

```{r}
y3 <- c(1, NA, NA, NA, NA, NA, NA, NA)
both_na(xtest, y3)
```

The issue is that the function is recycling the shorter vector to match the longer vector.

```{r}
3 != 2
```

```{r}
length(xtest)
```

```{r}
length(y2)
```

Now lets build a both_na2 function which will check for matching vector length

```{r}
both_na2 <- function(vec1, vec2) {
  
  if(length(vec1) != length(vec2)){
    stop("Hey dungus, these vectors aren't the same length!")
  }
   
  sum(is.na(vec1) & is.na(vec2))
}
```

```{r}
# both_na2(xtest,y3)
```
An error is generated

```{r}
which(c(F,F,T,F,T))

#which(is.na(c(1, 2, NA, 4)))
```


```{r}
x <- c(1, 2, NA, 3, NA)
y <- c(NA, 3, NA, 3, 4)

both_na3(x,y)
```

## Intersect Function: Looking at genes

```{r}
df1
```

```{r}
df2
```


Make things simpler


```{r}
x <- df1$IDs
y <- df2$IDs
x
y
```


```{r}
intersect(x,y)
```

```{r}
x %in% y
```

z
```{r}
which(x %in% y)
```

```{r}
x[x %in% y]
```

```{r}

```

```{r}
cbind(x[x %in% y],
      y[y %in% x] )

```

Using Code -> Extract Function shortcut in toolbar at top, we turned this code into a function

```{r}
gene_intersect <- function(x, y) {
  cbind( x[ x %in% y ], 
         y[ y %in% x ] )
}
```

```{r}
gene_intersect(df1$IDs, df2$IDs)
```


```{r}
gene_intersect2(df1,df2)
```

`

```{r}
gene_intersect3(df1,df2)
```

```{r}
gene_intersect4(df1,df2)
```







